How Short Lived is the Population III Epoch?

### 2010-09-08 ms.tex
Added references to halo and dSph metallicities.  Removed references to metal cooling.

\author{John H. Wise\altaffilmark{1,2}, 
  Michael L. Norman\altaffilmark{3},
  Tom Abel\altaffilmark{4},
  Britton D. Smith \altaffilmark{5}}

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton
  University, Peyton Hall, Ivy Lane, Princeton, NJ 08544}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Center for Astrophysics and Space Sciences,
  University of California at San Diego, La Jolla, CA 92093}
\altaffiltext{4}{Kavli Institute for Particle Astrophysics and
  Cosmology, Stanford University, Menlo Park, CA 94025}
\altaffiltext{5}{Michigan State University}
\email{jwise@astro.princeton.edu}

@@

The first (Pop III) stars are metal-free and have a large
characteristic mass and suppressed fragmentation in its protostellar
collapse \citep{Abel00, ABN02, Bromm02_P3, Yoshida03, OShea07a}.  A
fraction of these stars enrich the surrounding intergalactic medium
(IGM) when they go supernova, which can happen in stars $\lsim40~\Ms$
in Type II supernovae (SNe) or in stars roughly between 140 \Ms~and
260 \Ms~in pair-instability SNe \citep[PISNe;][]{2002ApJ...567..532H}.
The host halo and the neighboring halos are then enriched with this
ejecta.  There exists a critical metallicity that is $\sim 10^{-6}
Z_\odot$ if dust cooling is efficient \citep{Omukai05,
  Schneider06_Frag, clark08} and $\sim 10^{-3.5} Z_\odot$ otherwise
\citep{Bromm01, 2009ApJ...691..441S}, where the gas can cool rapidly,
lowering its Jeans mass.  An intermediate characteristic mass of
$\sim10~\Ms$ can be occur if the gas cooling is suppressed to the
cosmic microwave background (CMB) temperature \citep{Larson98,
  Tumlinson07_IMF, 2009ApJ...691..441S}.  The resulting Population II
star cluster will thus have a lower characteristic stellar mass than
its metal-free progenitors.  In the local universe, stars in the Milky
Way halo and dwarf spheroidal (dSph) galaxies may be connected to
these first stellar clusters.  In the halo of the Milky Way, the most
metal-poor stars have a metallicity floor of [Z/H] = ???  \citep{ref},
and in dwarf spheroidal (dSph) galaxies this floor is a factor of 10
higher \citep{refs}.

@@

Numerical simulations are useful to detangle and study these
complexities and the transition from Pop III to II stars.  In this
Letter, we present the simulations that include both types of star
formation, its radiative feedback, and its SNe feedback.  We utilize
an initial mass function (IMF) for individual Pop III stars, based on
the latest simulation results.  The methods used here incorporate and
link together recent results from metal-enriched and metal-free star
formation, the critical metallicity, pair-instability supernovae, and
fine-structure metal line cooling.  Furthermore we spatially
distinguish metal enrichment from Pop II and Pop III stars.  This is
the first time that all of these processes have been interconnected in
a simulation that resolves minihalos to study the transition to Pop II
star formation self-consistently.  We describe our simulation setup
and our adopted star formation models in \S \ref{sec:setup} and
present our results in \S \ref{sec:results}.  We discuss the
implications of our findings and conclude in the last section.

@@

We peform two simulations with the same initial conditions.  They
differ only in that one considers primordial cooling, and the other
considers additional cooling in metal-enriched gas.  For primordial
cooling, we use a non-equilbrium chemistry solver with 9 species of
hydrogen, helium, and molecular hydrogen.  For metal cooling, we use a
cooling curve that is calculated from CLOUDY \citep{CLOUDY} and using
the method of \citet{2008MNRAS.385.1443S}.  It is tabulated based on
density, temperature, electron fraction, and metallicity.

@@

For Pop III stars, we use the same star formation model as
\citet{Abel07} and \citet{Wise08_Gal} where each star particle
represents a single star.  The critical overdensity in which a star
forms is $5 \times 10^5$.  Instead of using a fixed stellar mass, we
randomly sampled from an IMF with a functional form of
%
\begin{equation}
\label{eqn:imf}
f(M)dM = M^{-1.3} \exp\left[-\left(\frac{M_{\rm char}}{M}\right)^{1.6}\right]
\end{equation}
to determine the stellar mass.  Above $M_{\rm char}$, it behaves as a
Salpeter IMF but is exponentially cutoff below that mass
\citep{Chabrier03, Clark09}.  For reproducibility, we initialize both
simulations with the same random seed and record the number of times
the random number generator \citep[Mersenne twister;][]{ref} has been
called for use when restarting the simulations.  Although the Pop III
star formation history (SFH) is identical for a given $N$ stars, the
Pop III SFH for a given halo can be different between the two
simulations.  This happens because the order in which halos form Pop
III stars can be altered from enhanced star formation and feedback
from nearby Pop II star clusters, as we will demonstrate in this
Letter.

We treat Pop II star formation with the same prescription as
\citet{Wise09}, which is a modified version of the widely used
\citet{Cen92} method but accounts for the fact that the molecular
clouds are resolved.  The critical overdensity is the same as the Pop
III star formation model.  In each star-forming region, a fraction
$c_* = 0.07$ of the cold gas ($T < 10^3$ K) is removed from the grid
and deposited into the star particle.  Each star particle lives for 20
Myr, the lifetime of an OB star.  These stars generate the majority of
the ionizing radiation and SNe feedback in stellar clusters, thus we
ignore lower mass stars.  We discuss our treatment of stellar feedback
next.

@@

>>>>

\author{John H. Wise\altaffilmark{1,2}, 
  Michael L. Norman\altaffilmark{3},
  Tom Abel\altaffilmark{4},
  Matthew J. Turk\altaffilmark{3}}

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton
  University, Peyton Hall, Ivy Lane, Princeton, NJ 08544}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Center for Astrophysics and Space Sciences,
  University of California at San Diego, La Jolla, CA 92093}
\altaffiltext{4}{Kavli Institute for Particle Astrophysics and
  Cosmology, Stanford University, Menlo Park, CA 94025}
%\altaffiltext{5}{Michigan State University}
\email{jwise@astro.princeton.edu}

@@

The first (Pop III) stars are metal-free and have a large
characteristic mass and suppressed fragmentation in its protostellar
collapse \citep{Abel00, ABN02, Bromm02_P3, Yoshida03, OShea07a}.  A
fraction of these stars enrich the surrounding intergalactic medium
(IGM) when they go supernova, which can happen in stars $\lsim40~\Ms$
in Type II supernovae (SNe) or in stars roughly between 140 \Ms~and
260 \Ms~in pair-instability SNe \citep[PISNe;][]{2002ApJ...567..532H}.
The host halo and the neighboring halos are then enriched with this
ejecta.  There exists a critical metallicity that is $\sim 10^{-6}
Z_\odot$ if dust cooling is efficient \citep{Omukai05,
  Schneider06_Frag, clark08} and $\sim 10^{-3.5} Z_\odot$ otherwise
\citep{Bromm01, 2009ApJ...691..441S}, where the gas can cool rapidly,
lowering its Jeans mass.  An intermediate characteristic mass of
$\sim10~\Ms$ can be occur if the gas cooling is suppressed to the
cosmic microwave background (CMB) temperature \citep{Larson98,
  Tumlinson07_IMF, 2009ApJ...691..441S}.  The resulting Population II
star cluster will thus have a lower characteristic stellar mass than
its metal-free progenitors.  In the local universe, stars in the Milky
Way halo and dwarf spheroidal (dSph) galaxies may be connected to
these first stellar clusters.  In the halo of the Milky Way, the most
metal-poor stars have a metallicity floor of [Z/H] = -4
\citep{Beers05}, and in dwarf spheroidal (dSph) galaxies this floor is
a factor of 10 higher \citep{Tolstoy04, Helmi06}.  \textbf{Check for
  more recent references.}

@@

Numerical simulations are useful to detangle and study these
complexities and the transition from Pop III to II stars.  In this
Letter, we present the simulations that include both types of star
formation, its radiative feedback, and its SNe feedback.  We utilize
an initial mass function (IMF) for individual Pop III stars, based on
the latest simulation results.  The methods used here incorporate and
link together recent results from metal-enriched and metal-free star
formation, the critical metallicity, and pair-instability supernovae.
Furthermore we spatially distinguish metal enrichment from Pop II and
Pop III stars.  This is the first time that all of these processes
have been interconnected in a simulation that resolves minihalos to
study the transition to Pop II star formation self-consistently.  We
describe our simulation setup and our adopted star formation models in
\S \ref{sec:setup} and present our results in \S \ref{sec:results}.
We discuss the implications of our findings and conclude in the last
section.

@@

We use a non-equilbrium chemistry solver with 9 species of hydrogen,
helium, and molecular hydrogen \citep{Abel97}.  We will follow-up this
study with one that considers radiative cooling from metal-enriched
gas, using rates that are calculated from CLOUDY \citep{CLOUDY} and
the method of \citet{2008MNRAS.385.1443S}.

% We peform two simulations with the same initial conditions.  They
% differ only in that one considers primordial cooling, and the other
% considers additional cooling in metal-enriched gas.  For primordial
% cooling, we use a non-equilbrium chemistry solver with 9 species of
% hydrogen, helium, and molecular hydrogen.  For metal cooling, we use a
% cooling curve that is calculated from CLOUDY \citep{CLOUDY} and using
% the method of \citet{2008MNRAS.385.1443S}.  It is tabulated based on
% density, temperature, electron fraction, and metallicity.

@@

For Pop III stars, we use the same star formation model as
\citet{Abel07} and \citet{Wise08_Gal} where each star particle
represents a single star.  The critical overdensity in which a star
forms is $5 \times 10^5$.  Instead of using a fixed stellar mass, we
randomly sampled from an IMF with a functional form of
%
\begin{equation}
\label{eqn:imf}
f(M)dM = M^{-1.3} \exp\left[-\left(\frac{M_{\rm char}}{M}\right)^{1.6}\right]
\end{equation}
to determine the stellar mass.  Above $M_{\rm char}$, it behaves as a
Salpeter IMF but is exponentially cutoff below that mass
\citep{Chabrier03, Clark09}.  For reproducibility, we initialize both
simulations with the same random seed and record the number of times
the random number generator \citep[Mersenne twister;][]{ref} has been
called for use when restarting the simulations.

%%% FOR THE COMPARISON BETWEEN RUNS WITH AND WITHOUT METAL COOLING %%%
%
% Although the Pop III star formation history (SFH) is identical for a
% given $N$ stars, the Pop III SFH for a given halo can be different
% between the two simulations.  This happens because the order in which
% halos form Pop III stars can be altered from enhanced star formation
% and feedback from nearby Pop II star clusters, as we will demonstrate
% in this Letter.

We treat Pop II star formation with the same prescription as
\citet{Wise09}, which is a modified version of the widely used
\citet{Cen92} method but accounts for the fact that the molecular
clouds are resolved.  The critical overdensity is the same as the Pop
III star formation model.  In each star-forming region, a fraction
$c_* = 0.07$ of the cold gas ($T < 10^3$ K) is removed from the grid
and deposited into the star particle.  Each star particle lives for 20
Myr, the maximum lifetime of an OB star.  These stars generate the
majority of the ionizing radiation and SNe feedback in stellar
clusters, thus we ignore lower mass stars.  We discuss our treatment
of stellar feedback next.

@@

\subsection{Metal enrichment history}

\subsection{Population III star formation history}

\subsection{Population II stellar populations}
